---
title: "Bike Sharing Demand"
author: "Akhil kumar"
date: "Tuesday, December 02, 2014"
output: html_document
---
## Introduction.
### Forecast use of a city bikeshare system

Bike sharing systems are a means of renting bicycles where the process of obtaining membership, rental, and bike return is automated via a network of kiosk locations throughout a city. Using these systems, people are able rent a bike from a one location and return it to a different place on an as-needed basis. Currently, there are over 500 bike-sharing programs around the world.

The data generated by these systems makes them attractive for researchers because the duration of travel, departure location, arrival location, and time elapsed is explicitly recorded. Bike sharing systems therefore function as a sensor network, which can be used for studying mobility in a city. In this competition, participants are asked to combine historical usage patterns with weather data in order to forecast bike rental demand in the Capital Bikeshare program in Washington, D.C.

The data is hosted on the UCI Machine Learning Repository <a http://archive.ics.uci.edu/ml/datasets/Bike+Sharing+Dataset>here<a/>.

Lets start digging into the data by loading the data.
```{r echo=TRUE,warning=FALSE,message=FALSE}
library(caret)
set.seed(100)
train <- read.csv("train.csv", header = T)
test <- read.csv("test.csv", header = T)
str(train)
```

As we can see the variables like weather,holiday,workingday & season are of type "int". They are binary variables, so we need to transform them into factors.
```{r echo=TRUE}
train$weather <- factor(train$weather)
train$holiday <- factor(train$holiday)
train$workingday <- factor(train$workingday)
train$season <- factor(train$season)
```
The same goes for testing.
```{r echo=TRUE}
test$weather <- factor(test$weather)
test$holiday <- factor(test$holiday)
test$workingday <- factor(test$workingday)
test$season <- factor(test$season)
```

I liked brandon's approach of creating new features, so followed his approach. Here's the link <a 
https://github.com/namebrandon/kaggle-bike-sharing/blob/master/kaggle-bike-challenge.R> here <a/>.
```{r, echo=TRUE}
#create time column by stripping out timestamp
train$time <- substring(train$datetime,12,20)
test$time <- substring(test$datetime,12,20)

#factorize new timestamp column
train$time <- factor(train$time)
test$time <- factor(test$time)
#create day of week column
train$day <- weekdays(as.Date(train$datetime))
train$day <- as.factor(train$day)
test$day <- weekdays(as.Date(test$datetime))
test$day <- as.factor(test$day)
#create Sunday variable
train$sunday[train$day == "Sunday"] <- "1"
train$sunday[train$day != "1"] <- "0"
test$sunday[test$day == "Sunday"] <- "1"
test$sunday[test$day != "1"] <- "0"
#convert to factor
train$sunday <- as.factor(train$sunday)
test$sunday <- as.factor(test$sunday)
#convert time and create $hour as integer to evaluate
train$hour<- as.numeric(substr(train$time,1,2))
test$hour<- as.numeric(substr(test$time,1,2))
train$daypart <- "4"
test$daypart <- "4"
#4AM - 9AM = 1
train$daypart[(train$hour < 10) & (train$hour > 3)] <- 1
test$daypart[(test$hour < 10) & (test$hour > 3)] <- 1
#10AM - 3PM = 2
train$daypart[(train$hour < 16) & (train$hour > 9)] <- 2
test$daypart[(test$hour < 16) & (test$hour > 9)] <- 2
#4PM - 9PM = 3
train$daypart[(train$hour < 22) & (train$hour > 15)] <- 3
test$daypart[(test$hour < 22) & (test$hour > 15)] <- 3
#convert daypart to factor
train$daypart <- as.factor(train$daypart)
test$daypart <- as.factor(test$daypart)
#convert hour back to factor
train$hour <- as.factor(train$hour)
test$hour <- as.factor(test$hour)
```
Now lets split the training data for training and validation.
```{r echo=TRUE}
split <- createDataPartition(train$count, p=0.8, list=F, times=1)
train_v <- train[split, ]
test_v <- train[-split, ]
```
Now lets build a model.
```{r echo=TRUE,cache=TRUE,warning=FALSE,message=FALSE}
library(party)
formula <- count ~ season + holiday + workingday + weather + temp + atemp + humidity + hour + daypart + sunday
ctree.model <- ctree(formula, data = train_v)
ctree_pred <- predict(ctree.model, newdata = test_v)
sse = sum((test_v$count - ctree_pred)^2)
sst = sum((test_v$count - mean(train_v$count))^2)
1-sse/sst # in-sample accuracy
```
lets now improve the model using different parametrs.
```{r echo=TRUE,warning=FALSE,message=FALSE}
library(party)
ctree.model1 <- ctree(formula, data=train_v, controls = ctree_control(teststat = "quad",
                     testtype = c("Univariate"),minsplit = 9, minbucket = 3,
                     mincriterion = 0.84))
ctree_pred <- predict(ctree.model1, newdata = test_v)
sse = sum((test_v$count - ctree_pred)^2)
sst = sum((test_v$count - mean(train_v$count))^2)
1-sse/sst # in-sample accuracy
```


